pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-05-08 bmc_article.tex

We next investigated the impact of filtering according to six methods with varying stringency; three methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}, see Methods for more details). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datsets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) often leads to high clustering performance (e.g. simMix and Zhengmix4 datasets) and always retain most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). This suggests that a mild ('default') filtering is a very good trade-off between clustering accuracy and subpopulations preservation.  Finally, the doublet removal step in conjunction with filtering  resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely (e.g. Zheng datasets). On the Smart-Seq and simulated datasets however, the removal of doublets had little to no impact on any of the two measures. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{SCTransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and SCnorm \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{SCTransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{SCTransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development of \textit{SCTransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 12). Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{SCTransform}. Surprisingly, however, regressing out the covariates tended to increase the association with it.

@@

\subsection*{Application of _pipeComp_ to a different context}

To illustrate the applicability of \textit{pipeComp} to a completely different setting, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the generalized linear model for DEA might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages implementing methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods (the package's `pipeComp\_dea' vignette details the creation process). We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details). As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 29A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 29A). Including a SVA-based step increased sensitivity across all datasets, and particularly so in datasets with technical vectors of variation, and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 29-30). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). These results therefore suggest that these methods are robust and can be widely applied.

@@

\subsection*{Software and package versions}
Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0). The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).

@@

The three benchmark datasets are available in \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper/blob/master/svadea/datasets/}, along with all code used to generate and prepare them. Briefly:
\begin{itemize}
    \item For the \textit{ipsc} dataset, we took the quantification of the GSE79636 dataset (fairly heterogeneous) from the \textit{RNAontheBENCH} package \citep{germainRNAontheBENCHComputationalEmpirical2016}, and selected two random sets of 10 samples. No further variation was added, and log-foldchanges ranging from 0.25 to 1.25 were added to the counts of 300 random genes. All other genes were assumed to be negatives.
    \item For the \textit{seqc} dataset, we used data from the \textit{seqc} package \citep{ComprehensiveAssessment2014}. For both groups (mixtures C and D, which also contain two different spike-in mixes), we selected samples from the \texttt{ILM\_refseq\_gene\_AGR} and \texttt{ILM\_refseq\_gene\_CNL} batches, which have clear technical differences; we selected 5 samples per group with a weak correlation with the batch (3:2 vs 2:3). Since the true differences between mixtures are not entirely known, the analysis was performed on all genes but only the spike-ins were considered for benchmarking.
    \item For the \textit{simulation}, we simulated counts using the \textit{polyester} package \citep{frazeePolyesterSimulating2015} and the means/dispersions from the GSE79636 dataset (restricted to a single batch and biological group), with 500 DEGs (log-foldchanges ranging from 0.5 to 2) and two batch effects: 1) a technical batch partially correlated with the group (6:2) and affecting 1/3 of the genes, and 2) a linear vector of variation uncorrelated with the groups and affecting 1/3 of the genes.
\end{itemize}

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/normalization.pdf}}
    \fi
    \caption{\textbf{Evaluation of normalization procedures. A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. Grey squares indicate that the true number of clusters could not be reached with the corresponding method and dataset. `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping. Other evaluation metric heatmaps (panels \textbf{A,C}) map colors linearly to the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}
    \label{fig:norm}
\end{figure}

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/svadea.pdf}}
    \fi
    \caption{\textbf{Evaluation the impact of prior factor analysis on (bulk) RNAseq differential expression analysis using \textit{pipeComp}. A:} True positive rates (TPR) and False discovery rates (FDR) of various combinations of SVA-based methods (using a single surrogate variable) and DEA methods. \textbf{B:} TPR and FDR of the different SVA-based methods (averaged across the different DEA methods) using increasing numbers of variables (`be' and `leek` refer to the two methods implemented in the \textit{sva} package for estimating the number of surrogate variables). In \textbf{A-B}, the x-axis is square-root transformed to improve readability, and three dots for each combination of methods refer to the nominal FDR thresholds of 0.01, 0.05, and 0.1. \textbf{C:} Correlation of estimated logFC with expected ones, as well as TPR and FDR (at nominal FDR threshold of 0.05) of the different methods with increasing numbers of variables. As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}
    \label{fig:svadea}
\end{figure}

>>>>

We next investigated the impact of filtering according to several methods (see Methods for more details): four methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12A). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datsets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) provided a good trade-off, often leading to high clustering performance (e.g. simMix and Zhengmix4 datasets) while retaining most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). We found very little overlap between the cells excluded by doublet removal and those excluded by MAD-based filtering (Supplementary Figure 12B). Of note, doublet removal in conjunction with filtering resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely given the experimental design (i.e. sorted datasets such as the Zheng datasets). Indeed, the datasets where doublet removal did not have an impact (e.g. Koh, Kumar, and the simulations) are those in which heterotypic doublets are not expected. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{SCTransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}, and \textit{SCnorm} \cite{bacherSCnorm2017} (\textit{scVI} \cite{lopezDeep2018} was included separately on a subset of parameter combinations due to high running times). In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{SCTransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function (and \textit{SCTransform}) also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development of \textit{SCTransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:norm}A and Supplementary Figure 14). Scaling tended to remove much of the correlation with these features, and while most methods were able to remove most of the effect, the correlation was lowest with \textit{SCTransform}. Surprisingly, however, regressing out the covariates tended to increase the association with it.

@@

\subsection*{Application of \textif{pipeComp} to a different context}

To illustrate the applicability of \textit{pipeComp} to a completely different setting, we benchmarked the impact of surrogate variable analysis (SVA) -based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva}\citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq}\citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three step: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citet{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A **simulation** of 8 vs 8 samples with two technical vectors of variation (see methods for more details). As previously reported, we found that more stringent filtering improved DEA (Supplementary Figure 29A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 29A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 29-30). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives, and the FDR nevertheless remained acceptable with 1-3 variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

\subsection*{Software and package versions}
Analyses were performed in R 3.6.0 (Bioconductor 3.9) and the following packages were installed from GitHub repositories: \textit{Seurat} (version 3.0.0), \textit{SCTransform} (0.2.0), \textit{DoubletFinder} (2.0.1), \textit{scDblFinder} (1.1.4), \textit{scds} (1.0.0), \textit{SAVERX} (1.0.0), \textit{scImpute} (0.0.9), \textit{ALRA} (commit 7636de8), \textit{DCA} (0.2.2), \textit{DrImpute} (1.2), \textit{ENHANCE} (R version, commit 1571696), \textit{SAVERX} (1.0.0). The glmPCA was performed using the \texti{glmpca} package (The code for the glmPCA was obtained from \url{https://github.com/willtownes/scrna2019} (commit 1ddcc30ebb95d083a685f12fe81d35dd1b0cb1b2).

@@

The three benchmark datasets are available in \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper/blob/master/svadea/datasets/}, along with all code used to generate and prepare them. Briefly:
\begin{itemize}
    \item For the \textit{ipsc} dataset, we took the quantification of the GSE79636 dataset (fairly heterogeneous) from the \textit{iPSCpower} package \citep{germainTamingHumanGenetic2017}, and selected two random sets of 10 samples. No further variation was added, and log-foldchanges ranging from 0.25 to 1.25 were added to the counts of 300 random genes. All other genes were assumed to be negatives.
    \item For the \textit{seqc} dataset, we used data from the \textit{seqc} package \citep{ComprehensiveAssessment2014}. For both groups (mixtures C and D, which also contain two different spike-in mixes), we selected samples from the \texttt{ILM\_refseq\_gene\_AGR} and \texttt{ILM\_refseq\_gene\_CNL} batches, which have clear technical differences; we selected 5 samples per group with a weak correlation with the batch (3:2 vs 2:3). Since the true differences between mixtures are not entirely known, the analysis was performed on all genes but only the spike-ins were considered for benchmarking.
    \item For the \textit{simulation}, we simulated counts using the \textit{polyester} package \citep{frazeePolyesterSimulating2015} and the means/dispersions from the GSE79636 dataset (restricted to a single batch and biological group), with 500 DEGs (log-foldchanges ranging from 0.5 to 2) and two batch effects: 1) a technical batch partially correlated with the group (6:2) and affecting 1/3 of the genes, and 2) a linear vector of variation uncorrelated with the groups and affecting 1/3 of the genes.
\end{itemize}

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/normalization.pdf}}
    \fi
    \caption{\textbf{Evaluation of normalization procedures. A:} Variance in PC1 explained by library size (left) and detection rate (right) after accounting for subpopulation differences. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by mutual information (MI), Adjusted Rand Index (ARI) and ARI at the true number of cluster. Grey squares indicate that the true number of clusters could not be reached with the corresponding method and dataset. `feat\_regress', `mt\_regress' and `feat\_mt\_regress' respectively stand for the regressing out of the number of features, the proportion of mitochondrial reads, or both during scaling. Throughout this paper, silhouette plots (panel \textbf{B}) square-root transform values for color mapping. Other evaluation metric heatmaps (e.g. panel \textbf{C} here) map colors linearly to the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median. Using this mapping, the same \textit{difference} in color has the same meaning across datasets, and the color-scale is not primarily capturing baseline differences between datasets. The numbers printed in the cells represent the raw (i.e. unscaled) metric values.}
    \label{fig:norm}
\end{figure}

@@

\begin{figure}
    \centering
    \ifshowfigs
    \includegraphics[width=\textwidth, keepaspectratio]{{main_figures/svadea.pdf}}
    \fi
    \caption{\textbf{Evaluation of the impact of surrogate variable analysis (SVA) on (bulk) RNAseq differential expression analysis using \textit{pipeComp}.} \textit{RUVr} and \textit{RUVs} refer to the two \textit{RUVSeq} functions by the same name; \textit{svaseq} refers to the function of the same name in \textit{sva}, while \textit{vstsva} refers to the original \textit{sva} method applied no variance-stabilized data. \textbf{A:} True positive rates (TPR) and False discovery rates (FDR) of various combinations of SVA-based methods (using a single surrogate variable) and DEA methods. \textbf{B:} TPR and FDR of the different SVA-based methods (averaged across the different DEA methods) using increasing numbers of variables (`be' and `leek` refer to the two methods implemented in the \textit{sva} package for estimating the number of surrogate variables). In \textbf{A-B}, the x-axis is square-root transformed to improve readability, and three dots for each combination of methods refer to the nominal FDR thresholds of 0.01, 0.05, and 0.1. \textbf{C:} Correlation of estimated logFC with expected ones, as well as TPR and FDR (at nominal FDR threshold of 0.05) of the different methods with increasing numbers of variables. As in \ref{fig:norm}C, the color mapping tracks the number of (matrix-wise) Median Absolute Deviations from the (column-wise) median, while the printed numbers represent the raw metric values.}
    \label{fig:svadea}
\end{figure}
